Variance reduction for MCMC methods via martingale representations

### 2020-02-20 aap/Revision/vr-mcmc-revision-new.tex
1st theorem proof update

%For a function $f\colon \rset^d \to \rset$ we define
%$$
%f\left(X^x_{p}\right) =%f\circ G_{p}(x,\sqrt{\gamma}Z_{1},\ldots,\sqrt{\gamma}Z_{p}).
%f\circ G_{p}(x,\xi_{1},\ldots,\xi_{p})
%$$
\begin{thm}\label{prop:29032018a1}
Let \((X^x_{p})_{p\geq 0}\) be a Markov chain of the form \eqref{eq:ula_new}. Then, for $p>j$ and for all Borel functions $f: \rset^{d}\to \rset$ such that $\mathsf{E}\left[\left |f(X^x_{p})\right|^{2}\right]<\infty$, the following representation holds in \(L^2(\mathsf{P})\)
\begin{eqnarray}
\label{eq:mart_repr}
f(X^x_{p})=\mathsf{E}\left[\left.f(X^x_{p})\right|\mathcal G_{j}\right]+\sum_{k=1}^{\infty}\sum_{l=j+1}^{p}a_{p,l,k}(X^x_{l-1})\phi_k\left(\xi_{l}\right),
\end{eqnarray}
where for $y \in \rset^d$
\begin{eqnarray}
\label{eq:coeff_mart}
%a_{p,l,k}(x)=\mathsf{E}\left[\left.f(X_{p})\phi_k\left(\xi_{l}\right)\right|X_{l-1}=x\right], \quad p\geq l, \quad k\in \mathbb{N}.
a_{p,l,k}(y)=\mathsf{E}\left[f(X^y_{l-1,p})\phi_k\left(\xi_{l}\right)\right], \quad p\geq l, \quad k\in \mathbb{N}
\end{eqnarray}
\end{thm}

\begin{proof}%[Proof of Theorem~\ref{prop:29032018a1}]
The expansion obviously holds for $p=1$ and $j=0$.
Indeed, since $\left(\phi_{k}\right)_{k \geq 0}$ is a complete orthonormal system in \(L^2(\mathbb{R}^d, P_{\xi})\), it holds in \(L^2(\mathsf{P})\) that
\[
f(X^x_{1})=\PE[f(X^x_{1})]+\sum_{k\geq1}a_{1,1,k}(x)\phi_{k}(\xi_{1})
\]
with $a_{1,1,k}(x)=\PE[f(X_{1}^x)\phi_{k}(\xi_{1})]$.
Suppose that (\ref{eq:mart_repr})
holds for $p=q$, all $j<q$, and all Borel-measurable functions
$f$ with $\mathsf{E}\left[|f(X^x_{q})|^2\right]<\infty$.
Let us prove it for $p=q+1$.
Given $f$ with $\mathsf{E}\left[|f(X^x_{q+1})|^2\right]<\infty$,
due to the orthonormality and completeness
of the system $\left(\phi_{k}\right)$, we get by
conditioning on $\mathcal{G}_{q}$,
\[
f(X^x_{q+1})=\mathsf{E}\left[\left.f(X^x_{q+1})\right|{\mathcal{G}}_q\right]+\sum_{k\geq1}\alpha_{q+1,q+1,k}\phi_{k}(\xi_{q+1}),
\]
where
\begin{equation*}
\alpha_{q+1,q+1,k}
=\mathsf{E}\left[\left.f(X^x_{q+1})\phi_{k}(\xi_{q+1})\right|\mathcal G_q\right].
\end{equation*}
By the Markov property of $(X^x_{l})$, it holds $\mathsf{E}[f(X^x_{q+1})|\mathcal{G}_q] = \mathsf{E}[f(X^x_{q+1})| X_q^x]$. Using Markov property together with conditioning on $\mathcal{G}_{q+1}$, we obtain
\[
\alpha_{q+1,q+1,k}=\mathsf{E}\left[\left.f(X^x_{q+1})\phi_{k}(\xi_{q+1})\right|X^x_{q}\right]
=a_{q+1,q+1,k}(X^x_{q})
\]
where for $y \in \rset^d$
\[
a_{q+1,q+1,k}(y) = \mathsf{E}\left[f(X^y_{q,q+1})\phi_k\left(\xi_{q+1}\right)\right]
\]

>>>>

\begin{thm}\label{prop:29032018a1}
Let \((X^x_{p})_{p\geq 0}\) be a Markov chain of the form \eqref{eq:ula_new}. Then, for $p>j$ and for all Borel functions $f: \rset^{d}\to \rset$ such that $\mathsf{E}\left[\left |f(X^x_{p})\right|^{2}\right]<\infty$, the following representation holds in \(L^2(\mathsf{P})\)
\begin{eqnarray}
\label{eq:mart_repr}
f(X^x_{p})=\mathsf{E}\left[\left.f(X^x_{p})\right|\mathcal G_{j}\right]+\sum_{k=1}^{\infty}\sum_{l=j+1}^{p}a_{p,l,k}(X^x_{l-1})\phi_k\left(\xi_{l}\right),
\end{eqnarray}
where for $y \in \rset^d$
\begin{eqnarray}
\label{eq:coeff_mart}
%a_{p,l,k}(x)=\mathsf{E}\left[\left.f(X_{p})\phi_k\left(\xi_{l}\right)\right|X_{l-1}=x\right], \quad p\geq l, \quad k\in \mathbb{N}.
a_{p,l,k}(y)=\mathsf{E}\left[f(X^y_{l-1,p})\phi_k\left(\xi_{l}\right)\right], \quad p\geq l, \quad k\in \mathbb{N}
\end{eqnarray}
\end{thm}

\begin{proof}%[Proof of Theorem~\ref{prop:29032018a1}]
The expansion obviously holds for $p=1$ and $j=0$.
Indeed, since $\left(\phi_{k}\right)_{k \geq 0}$ is a complete orthonormal system in \(L^2(\mathbb{R}^d, P_{\xi})\), it holds in \(L^2(\mathsf{P})\) that
\[
f(X^x_{1})=\PE[f(X^x_{1})]+\sum_{k\geq1}a_{1,1,k}(x)\phi_{k}(\xi_{1})
\]
with $a_{1,1,k}(x)=\PE[f(X_{1}^x)\phi_{k}(\xi_{1})]$.
Assume now that (\ref{eq:mart_repr})
holds for $p=q$, all $j<q$, and Borel-measurable functions
$f$ with $\mathsf{E}\left[|f(X^x_{q})|^2\right]<\infty$.
Let us prove that the induction assumption holds for $p=q+1$.
Given $f$ with $\mathsf{E}\left[|f(X^x_{q+1})|^2\right]<\infty$,
due to the orthonormality and completeness
of the system $\left(\phi_{k}\right)$, we get that for any $y \in \rset^d$ it holds in \(L^2(\mathsf{P})\) 
\[
f(X^y_{q,q+1}) = \mathsf{E} f(X^y_{q,q+1}) + \sum\limits_{k=1}^{\infty}\mathsf{E}[f(X^y_{q,q+1})\phi_{k}(\xi_{q+1})]\phi_k(\xi_{q+1})
\]
which means that
\begin{equation}
\label{eq:parseval}
\lim_{n \rightarrow \infty}\mathsf{E}[|f(X^y_{q,q+1}) - \mathsf{E} f(X^y_{q,q+1}) - \sum\limits_{k=1}^{n}\mathsf{E}[f(X^y_{q,q+1})\phi_{k}(\xi_{q+1})]\phi_k(\xi_{q+1})|^2] = 0
\end{equation}
Let us define sequence $\psi_{q+1,n}(y) = \mathsf{E}[|f(X^y_{q,q+1}) - \mathsf{E} f(X^y_{q,q+1})|^2] - \sum\limits_{k=1}^{n}\mathsf{E}^2[f(X^y_{q,q+1})\phi_k(\xi_{q+1})]$. Note that $\psi_{q+1,n}(y)$ coincides with the left side of \ref{eq:parseval}, and due to Parseval identity $\Psi_{q+1,n}(y) \rightarrow 0, n \rightarrow \infty$ for any $y \in \rset^d$. Note that $\psi_n(y) < \mathsf{E}[|f(X^y_{q,q+1}) - \mathsf{E} f(X^y_{q,q+1})|^2] \leq \frac{\|f\|^2_{\infty}}{4}$. Hence, by Lebesgue dominated convergence theorem, $\mathsf{E}\psi_n(X_q^x)^2 \rightarrow 0$ and since $\mathsf{E}[f(X^y_{q,q+1})\phi_{k}(\xi_{q+1})]$ is a version of conditional expectation $\mathsf{E}[f(X^x_{q+1}|X^x_q = y]$, it holds in \(L^2(\mathsf{P})\)
\begin{equation}
\label{eq:1-step-decomp}
f(X^x_{q+1}) = \mathsf{E}[f(X^x_{q+1})|\mathcal{G}_k] + \sum\limits_{k=1}^{\infty}a_{q+1,q+1,k}(X^x_{q})\phi_{k}(\xi_{q+1})
\end{equation}
where
\[
a_{q+1,q+1,k}(y) = \mathsf{E}\left[f(X^y_{q,q+1})\phi_k\left(\xi_{q+1}\right)\right]
\]

%we get by
%conditioning on $\mathcal{G}_{q}$,
%\[
%f(X^x_{q+1})=\mathsf{E}\left[\left.f(X^x_{q+1})\right|{\mathcal{G}}_q\right]+\sum_{k\geq1}\alpha_{q+1,q+1,k}\phi_{k}(\xi_{q+1}),
%\]
%where
%\begin{equation*}
%\alpha_{q+1,q+1,k}
%=\mathsf{E}\left[\left.f(X^x_{q+1})\phi_{k}(\xi_{q+1})\right|\mathcal G_q\right].
%\end{equation*}
%By the Markov property of $(X^x_{l})$, it holds $\mathsf{E}[f(X^x_{q+1})|\mathcal{G}_q] = \mathsf{E}[f(X^x_{q+1})| X_q^x]$. Using Markov property together with conditioning on $\mathcal{G}_{q+1}$, we obtain
%\[
%\alpha_{q+1,q+1,k}=\mathsf{E}\left[\left.f(X^x_{q+1})\phi_{k}(\xi_{q+1})\right|X^x_{q}\right]
%=a_{q+1,q+1,k}(X^x_{q})
%\]
%where for $y \in \rset^d$
%\[
%a_{q+1,q+1,k}(y) = \mathsf{E}\left[f(X^y_{q,q+1})\phi_k\left(\xi_{q+1}\right)\right]
%\]
